SILVR: Conditional diffusion model for molecule generation without additional training

### 2023-03-14 paper/manuscript.tex
Updates from Overleaf

\section{Methods}
The selective iterative latent variable refinement (SILVR) method makes use of the pre-trained equivariant diffusion model (EDM) as introduced by Hegeboom et al.~\cite{}. The core of the method is the introduction of an additional refinement within the denoising process during run time. The resulting SILVR model can then produce conditional samples without any conditional training. In this work, we show how SILVR can be applied to fragment-based drug design to produce molecules resembling a reference set of atomic coordinates.

The EDM by Hoodgeboom et al was trained on the 30 lowest energy conformations of 430,000 molecules from the GEOM dataset with explicit protons. A simplified schematic describing EDM architecture is shared in scheme X for more details on the model theory and training refer to Hegeboom et al.~\cite{}. This model strictly only considers atomic coordinates; all bond information is ignored. During training, the atomic coordinates (and their element) are passed through a forward diffusion process with iterative addition of Gaussian noise, where the extent of noise added at each step is defined by parameter $\beta$ (N.B. scheme X shows the diffusion process as a Markov chain, however in practice the state at time $t$ is determined as a direct transformation of the initial state). The diffusion process is eventually terminated when $t=T$, by which point all structure is lost and all coordinates follow a gaussian distribution. An equivariant graph neural network (EGNN) is then trained to predict the reverse process, predicting the previous state in the sequence given any state. At run time, the generative model is instantiated with a sample from a gaussian distribution and the series of denoising steps are applied resulting in a generated sample consisting of atomic coordinates resembling a drug-like molecule. The XYZ coordinates can then be interpreted using cheminformatics software to determine the molecular graph. As described, the EDM model is only trained to generate random molecules; hoodgeboom’s model does enable training given additional context, however this is outside the scope of this paper.

>>>>

\section{Methods}
The selective iterative latent variable refinement (SILVR) method makes use of the pre-trained equivariant diffusion model (EDM) as introduced by Hoogeboom \textit{et al.}~\cite{hoogeboom2022equivariant}. The core of the method is the introduction of an additional refinement within the denoising process during run time. The resulting SILVR model can then produce conditional samples without any conditional training. In this work, we show how SILVR can be applied to fragment-based drug design to produce molecules resembling a reference set of atomic coordinates.

The EDM by Hoodgeboom \textit{et al.} was trained on the 30 lowest energy conformations of 430,000 molecules from the GEOM dataset with explicit protons. A simplified schematic describing EDM architecture is shared in scheme X for more details on the model theory and training refer to Hoogeboom et al.~\cite{hoogeboom2022equivariant}. This model strictly only considers atomic coordinates; all bond information is ignored. During training, the atomic coordinates (and their element) are passed through a forward diffusion process with iterative addition of Gaussian noise, where the extent of noise added at each step is defined by parameter $\beta$ (N.B. scheme X shows the diffusion process as a Markov chain, however in practice the state at time $t$ is determined as a direct transformation of the initial state). The diffusion process is eventually terminated when $t=T$, by which point all structure is lost and all coordinates follow a gaussian distribution. An equivariant graph neural network (EGNN) is then trained to predict the reverse process, predicting the previous state in the sequence given any state. At run time, the generative model is instantiated with a sample from a gaussian distribution and the series of denoising steps are applied resulting in a generated sample consisting of atomic coordinates resembling a drug-like molecule. The XYZ coordinates can then be interpreted using cheminformatics software to determine the molecular graph. As described, the EDM model is only trained to generate random molecules; Hoodgeboom’s model does enable training given additional context, however this is outside the scope of this paper.
